A generalization of Plackett-Luce likelihoods to incorporate non-independence of competitors

### 2023-01-08 inst/plackett_luce_monster.Rmd

# Start simple

The simplest nontrivial case is three runners, $\left\lbrace
1,2,3\right\rbrace$ where runner 1 helps runner 2 [and that is the
only non PL interaction].  Note that this is _not_ a symmetric
interaction: 1 helps 2, but 2 does not help 1.  The model is that 1
helps 2 if and only if 1 has finished and 2 has not.

@@

231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$

@@

(Above, the probabilities clearly sum to 1).  Now the same but with
four runners $\left\lbrace 1,2,3\right\rbrace$, again with runner 1
helping runner 2.  Thus, in a situation where $p_1$ has finished and
$p_2$ is still running, we have $p_2\longrightarrow\lambda p_2$
[although the $\lambda$ cancels if 2 is the only competitor still
running, that is, if *all* the remaining competitors are helped]

@@

# Next simplest: 1 and 2 mutually support each other:

123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda 2}{\lambda p_2+p_3}$

# Four runners

>>>>

# Start simple: three runners

The simplest nontrivial case is three runners, $\left\lbrace
1,2,3\right\rbrace$ where runner 1 helps runner 2 with multiplier
$\lambda$ [and that is the only non PL interaction].  Note that this
is _not_ a symmetric interaction: 1 helps 2, but 2 does not help 1.
The model is that 1 helps 2 if and only if 1 has finished and 2 has
not.

@@

231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{p_1+p_3}$

@@

(Above, the probabilities clearly sum to 1).  Next simplest: 1 and 2
mutually support each other with $\lambda$:

Now we retain the mutual support between 1 and 2 but allow an
asymmetry: 1 helps 2 with $\lambda$ as before, but now in addition, 2
helps 1 with $\mu$:

213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\mu p_1}{\mu p_1+p_3}$

231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\mu p_1+p_3}$

Now how about 1 must beat 2.  He does not care whether he comes first
or second, only that he must beat 2.  Player 1 has sufficient energy
to win if he wants, but is subject to some sort of resource constraint
(money, perhaps) and wishes to expend only enough money to beat 2.  So
conventional P-L likelihoods are modified for 1, but only if 2 is
_still running_.

123: $\frac{\lambda p_1}{\lambda p_1+p_2+p_3}\cdot\frac{p_2}{p_2+p_3}$

132: $\frac{\lambda p_1}{\lambda p_1+p_2+p_3}\cdot\frac{p_3}{p_2+p_3}$

213: $\frac{p_2}{\lambda p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$

231: $\frac{p_2}{\lambda p_1+p_2+p_3}\cdot\frac{p_3}{p_1+p_3}$

312: $\frac{p_3}{\lambda p_1+p_2+p_3}\cdot\frac{\lambda p_1}{\lambda p_1+p_2}$

321: $\frac{p_3}{\lambda p_1+p_2+p_3}\cdot\frac{p_2}{\lambda p_1+p_2}$

## Four runners

Now we have four runners $\left\lbrace 1,2,3,4\right\rbrace$, again
with the only interaction being runner 1 helping runner 2.  Thus, in a
situation where $p_1$ has finished and $p_2$ is still running, we have
$p_2\longrightarrow\lambda p_2$ [although the $\lambda$ cancels if 2
is the only competitor still running, that is, if *all* the remaining
competitors are helped]

@@

Now 1 and 2 mutually support one another by $\lambda$.  Here I will
put the last term in, to show explicitly that the $\lambda$ term
remains in the expressions for numerator and denominator.

1234: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
p_2+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}\cdot\frac{p_4}{p_4}$

1243: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
p_2+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}\cdot\frac{p_3}{p_3}$

1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda
p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
p_2+p_4}\cdot\frac{p_4}{p_4}$

1342: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda
p_2+p_3+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}\cdot\frac{\lambda
p_2}{\lambda p_2}$

1423: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda
p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
p_2+p_3}\cdot\frac{p_3}{p_3}$

1432: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda
p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3}\cdot\frac{\lambda
p_2}{\lambda p_2}$

2134: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
p_1+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}\cdot\frac{p_4}{p_4}$

2143: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
p_1+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}\cdot\frac{p_3}{p_3}$

2314: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_3}{\lambda
p_1+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
p_1+p_4}\cdot\frac{p_4}{p_4}$

2341: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_3}{\lambda
p_1+p_3+p_4}\cdot\frac{p_4}{\lambda p_1+p_4}\cdot\frac{\lambda
p_1}{\lambda p_1}$
